---
title: "simple_linear_regression"
author: "Walter J.T.V"
date: "`r Sys.Date()`"
output: html_document
---

# Simple Linear Regression

2 tools for validating models (even if they are not lineal): 

- Residual analysis 
- Cross Validation

## 1. Dataset Loading

```{r}

load("Anscombe73raw.RData")
names(anscombe)
df <- anscombe
```

## 2. Dataset A

```{r}



model_a <- lm(YA ~ XA, data=df)
summary(model_a)
# The model is Ya = 3.0001 + 0.5001Xa + e_x
# A hypothesis testing of the true values of the parameters is performed with the summary
# 1. Ho: B1 = 0, H1: B1 != 0  
# 2. Ho: B2 = 0, H1: B2 != 0

# Also residual standard error S² = RSS/n-p with n-p degrees of freedom (p is the num of vars)
# Also multiple R-squared are presented (How many variability is explained by the model)
# The rest is explained by the residuals

# Finally a F-statistic and it's p-value are computed: F = MeanSquareRegression / MeanSquareResiduals
# This tells us how 


plot(df$XA, df$YA, pch=19, col="black")
lines(df$XA, fitted.values(model_a), col="red")
```

```{r}

plot(model_a, id.n=0)

# Residuals vs Fitted: We want to check that residuals are random and do not follow a pattern, otherwise it means that a lot of patterns and variance are left unexplained. WE WANT TO CHECK LINEARITY AND HOMOSCEDASTICITY

# If u-shape or curved patterns are met, then its either heteroscedastic or not linear

# Q-Q Residuals: Quantiles of residuals are plot to check if residuals are normally distributed

# Scale-Location: To check an assumption like homoscedasticity (Constant variance of errors)

```
## 3. Model B 
```{r}

model_b <- lm(YB ~ XB, data=df)
summary(model_b)

plot(df$XB, df$YB, col="black", pch=19)
lines(df$XB, model_b$fitted.values, col="red")
```
Looking at the Residual vs Fitted plot we can asses that residuals do not follow a random noise, they follow a pattern, then the model is not linear nor obey homocedasticity.

On scale location we can also see a weird pattern that points out that the variance of residuals is not constant.
```{r}

plot(model_b)

```


Lets create a new quadratic model B and get funky 

```{r}
model_b2 <- lm(YB ~ XB + I(XB^2), data=df)
summary(model_b2)
#The models fits perfectly the data

plot(model_b2)
#lines(df$XB, model_b2$fitted.values, col="red")
```

## 4. Model C

```{r}

model_c <- lm(YC ~ XC, data=df)
summary(model_c)

plot(df$XC, df$YC, col="darkgreen", main="Dataset C")
text(df$XC, df$YC, label=rownames(df), col="darkgreen", adj=-1)
lines(df$XC, model_c$fitted.values, col="blue")

```

Cook distance can be used to detect influential data (OUTLIERS)
```{r}
library(car)
boxplot(resid(model_c))
plot(model_c)
boxplot(cooks.distance(model_c))
influencePlot(model_c)

```

Lets erase the outlier and get funky:

```{r}

model_c2 <- lm(YC ~ XC, data=df[-3,])
summary(model_c2)

```

## 5. Model D 

```{r}

model_d <- lm(YD ~ XD, data=df)
summary(model_d)


plot(df$XD, df$YD, pch=19, col="blue", main="Dataset D")
lines(df$XD, model_d$fitted.values, col="red")
```

```{r}

plot(model_d)

cooks.distance(model_d)
hatvalues(model_d)

```
