---
title: "Introduction to R, RStudio"
author: "Lecturers FIB-MDS-SIM"
date: \today
output:
  word_document:
    toc: no
    toc_depth: '4'
  html_document:
    toc: no
    toc_depth: '4'
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
geometry: left=1.9cm,right=1.9cm,top=1.25cm,bottom=1.52cm
fontsize: 18pt
subtitle: 'Laboratori 0 - Basic commands'
classoption: a4paper
editor_options: 
  chunk_output_type: console
---

# Principal characteristics of R and its programming:
- Open-source.

- Highly Active Community. Functions and packages are personal creations from users. 

- Oriented to objects. 

- Extremely comprehensive. 

- It does not need a compiler to run code. 

- Direction to machine learning. 

- Compatibility with other Data Processing Technologies (Example: Use a spark cluster to process large datasets using R)

- R markdown to generate reports in any desired format.

- Operations directly on vectors, not too much looping.

- Data from APIs (and many other formats) can be easily pulled down.

# Initial commands and basic descriptive statistics:

How to citate R?

```{r}
citation()
```

Load package and data:

```{r}
# install.packages("car")
library(car)
data(Davis)
attributes(Davis)
str(Davis)
# Numeric Univariant Description
summary(Davis)
# Rows and columns of data.frame Davis
dim(Davis)
```

Graphical description:

```{r}
# Graphical Description of numeric data

# Histograms:
par(mfrow=c(2,2))
hist(Davis$weight)
hist(Davis$weight,10)
hist(Davis$weight,10,col="blue")
hist(Davis$weight,10,col=rainbow(10))

# Histogram with proportions:
par(mfrow=c(1,1))
hist(Davis$weight,freq=F)  # Proportions

# Dotplot:
par(mfrow=c(2,1))
stripchart(Davis$weight,method="stack")
stripchart(Davis$weight,method="stack",xlab="weight",pch=19,col=3,main="Dotplot Weight in Davis dataset")

# Boxplots (two ways):
par(mfrow=c(2,3))
boxplot(Davis$weight)
boxplot(Davis$weight,col="blue",horizontal = TRUE)
boxplot(Davis$weight,col="blue",horizontal = TRUE, pch=19,labels=Davis$weight)

Boxplot(Davis$weight)
Boxplot(Davis$weight,col="blue",main= "Weight in Davis dataset - row name Id")
Boxplot(Davis$weight,col="blue",main=" Boxplot Weight - Weight Label for Outliers",labels=Davis$weight)

#Barplots and pie charts:
par(mfrow=c(2,2))
barplot(table(Davis$sex))
barplot(table(Davis$sex),col=rainbow(2))
pie(table(Davis$sex))
pie(table(Davis$sex),col=rainbow(2))

```

Description of variable factors:

```{r}
table(Davis$sex)
margin.table(table(Davis$sex))
prop.table(table(Davis$sex))
```

Ask for information, arguments and outputs of a function:

```{r}
# ?boxplot
```

Other functions:


```{r}
# View(Davis)
head(Davis, n = 20) # n = 20 means  that the first 20 lines are printed in the R console

attach(Davis)
summary(weight)
detach(Davis)
# summary(weight) # Do not work

with(Davis,tapply(height,sex,summary))
summary(Davis$height)
with(Davis,Boxplot(height~sex,pch=19,col=rainbow(2),main="Height vs Sex"))
```

R markdown offers the possibility to structure the document in different header levels:

# Header 1

## Header 2

### Header 3

and simple text included into pargraphs. 

To enumerate:

- Enumeration 1

- Enumeration 2

- ...

It also allows the user to write in **bold** and in *italica*. 

In addition, equations in $\LaTeX$ can be added when you are writing such as $2 \cdot x = 6$ and appart from text:

$$2 \cdot x = 6$$
More elements which can be included are: 

- $\LaTeX$ tables

- Images

- Links

- Bibliography

- $\LaTeX$ matrices

- ...

More possibilities can be found visiting next reference:

[10 R markdown possibilities](https://towardsdatascience.com/ten-awesome-r-markdown-tricks-56ef6d41098)