---
title: "Practice_1"
author: "Walter J.T.V"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

## Dataset
We are going to use the prestige dataset which has 102 rows of dfferent professions and its associated infered prestige percieved by society

```{r}
rm(list=ls())
library(car)
prestige_df <- Prestige
head(prestige_df)
```

## Univariate outlier detection
Detection of univariate outliers using boxplots.
- Mild outliers: fall between the range (Q1 - 1.5IQR, Q3 + 1.5IQR)
- Extreme outliers: fall between the range (Q1 - 3IQR, Q3 + 3IQR)

The IQR (inter qualite range) is computed IQR = Q3 - Q1

```{r}
#Remove census column
prestige_df$census <- NULL
names(prestige_df)
numeric_prestige <- prestige_df[, c("education", "income", "women", "prestige")]
summary(numeric_prestige)

prestige_box <- boxplot(numeric_prestige$prestige) # No mild outliers and no severe outliers either

``` 

Prestige variable has no outliers, now lets check the average income

```{r}
boxplot(numeric_prestige$income, horizontal = T, col= "blue") # There are outliers
income_summary <- summary(numeric_prestige$income)

iqr <- income_summary[5] - income_summary[2]
upmild <- income_summary[5] + 1.5*iqr
upextreme <- income_summary[5] + 3*iqr

boxplot(numeric_prestige$income, horizontal=T, col="blue")
abline(v=upmild, col="orange")
abline(v=upextreme, col="red")

mild_outliers_ids <- which( (prestige_df$income > upmild) & (prestige_df$income < upextreme))

ext_outliers_ids <- which( (prestige_df$income > upextreme))

mild_outliers <- prestige_df[mild_outliers_ids, ]
ext_outliers <- prestige_df[ext_outliers_ids, ]

income_box <- Boxplot(numeric_prestige$income, col="green", id = list(labels=rownames(prestige_df)))
income_box

#Range can specify the Quartiles
```


## Multivariate outlier detection
Now we are going to define a distance measure in order to detect **Multivariate outliers** using an iterative algorithm. That is the Mahalanobis distance. 

Mahanalobis distances follows a chi-squared distribution, therefore the distance depends in the distance from the center of gravity 


NOTE: 

```{r}
  library(chemometrics)
  res.mout <- Moutlier(numeric_prestige, quantile=0.95)
  plot(x=res.mout$md, y=res.mout$rd, main="Mahalanobis: Classic vs Robust", xlab="Classic distance", ylab="Robust distance")
  abline(v=res.mout$cutoff, col="red")
  abline(h=res.mout$cutoff, col="red")
  text(res.mout$md, res.mout$rd, labels=rownames(numeric_prestige))
  
  mv_outliers <- which( (res.mout$md > res.mout$cutoff) & (res.mout$rd > res.mout$cutoff))
  
  res.mout$cutoff^2 == qchisq(0.95, 4)
```

## Missing Data Types

There are different kinds of missing data:

- Non-response (All variables are null in an observation)
- Missing values (One or two variables)
- Drop outs (Several missing variables in an observation)

And just removing missing data is bad for the model generalization, but there are 3 procedures:

- Removal of observations with missing data (Bias in results)
- Unconditional mean imputation (HORRIBLE)
- Regression imputation

Following up we can see an example of imputation using the mice library

```{r}

```
  
  



