---
title: "logistic_regression2"
author: "Walter J.T.V"
date: "2024-12-02"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Data loading 

```{r}
load("Elections_92.RData")

library(MASS)
library(car)
library(caret)
library(ModelMetrics)
library(cvAUC)
#library(DescTools)
library(FactoMineR)
library(AER)
library(effects)
library(lmtest)
```

# EDA 

```{r}
summary(elecc92)
```

# Individual data vs Grouped data

$$ Model_1: \eta = \beta_0 + \beta_1 * x_{age}$$
## Individual data
```{r}
# Canonical link function (default) is logit
m1 <- glm(pres ~ age, family=binomial, data=elecc92)
summary(m1)

# logodds; every extra year in age increases by .013
# odds: exp(0.013) -> each year odds are multiplied by 1.014 more or less
# The increment in the odds scale is 1.35% every year
exp(coef(m1))
100 * (exp(coef(m1)[2]) - 1)

#Approximation to a probability scale :) (0-1)
mean(elecc92$pres) * (1 - mean(elecc92$pres)) * coef(m1)[2]

#How to predict manually the linear predictor and probability of voting at a median age? 
med_age <- median(elecc92$age)
eta <- coef(m1)[1] + coef(m1)[2] * med_age; eta # eta is the value of the linear predictor
pprob <- exp(eta) / (1+exp(eta)); pprob

# Effect of 10 years of increase
coef(m1)[2] * 10
# Odds: they increase by 14.4% with respect to the median
exp(coef(m1)[2] * 10)
```

## Grouped data

$$ Model_{1a}: \eta = \beta_0 + \beta_1 * x_{age}$$
Asimptotical properties of grouped data does not hold????

Both models have different absolute deviances, (Null and model) but when compared they are relatively the same.

You cannot compare deviances of 2 different models using Akaike because they use 2 different data-sets, (grouped and individual)

```{r}
head(dfage)
m1a <- glm(cbind(ypos, yneg) ~ age, family=binomial, data=dfage)
summary(m1a)
coef(m1) # Estimates B0, B1 for grouped data

#Differences between deviance in M0, M1, M0A, M1A
m1$null.deviance - m1$deviance
m1a$null.deviance - m1a$deviance

#Deviance test
Anova(m1, test="LR")
Anova(m1a, test="LR")
```


# Modelling

## Split 

```{r}
# work (train) and test datasets
set.seed(1234) 
ids_work <- sample(1:nrow(elecc92), round(0.8 * nrow(elecc92)))
dfwork <- elecc92[ids_work, ]
dftest <- elecc92[-ids_work, ]
```

## Numeric variables

```{r}
m2 <- glm(pres ~ age + educ, family=binomial, data=dfwork)
summary(m2)
Anova(m2, test="LR")

residualPlots(m2)
marginalModelPlots(m2)
```


```{r}
m3 <- glm(pres ~ poly(age,2) + poly(educ, 3), family=binomial, data=dfwork)
summary(m3)
marginalModelPlots(m3)
#still the model is not that good

m4 <- glm(pres ~ poly(age,2) *c.edu, family=binomial, data=dfwork)
summary(m4)
marginalModelPlots(m4)

m5 <- glm(pres ~ c.age*c.edu, family=binomial, data=dfwork)
summary(m5)

AIC(m1, m2, m3, m4, m5) # M4 is pretty god

library(effects)
plot(allEffects(m4))
#We are using effects, therefore we have a parabola for factors

```

## Categorical variables

```{r}

summary(dfwork)

m6 <- glm(pres ~ poly(age,2)*c.edu + inter + close + sat, family=binomial, data=dfwork)
summary(m6)

#They are not equivalent (2 tests for the same)
anova(m4, m6, test="Chisq")  #Statistic = difference in deviance
waldtest(m4, m6, test="Chisq") #Statistic = chi-squared 

coef(m6) #Check sat: Yes factor if its equal to 0 or more than 1
linearHypothesis(m6, c("satSat-Yes=0", "closeClose-Yes=0"))


m6aux <- glm(pres ~ poly(age,2) * c.edu + inter, family=binomial, data=dfwork)
anova(m6aux, m6, test="Chisq")

#Weak version net effects
Anova(m6, test="LR")
```

## Party variable 

```{r}

summary(dfwork)
dfwork$party <- paste(dfwork$p, dfwork$arty)
head(dfwork)


m7 <- glm(pres ~ poly(age,2)*c.edu + inter + close + sat + party, family=binomial, data=dfwork)
summary(m7)
Anova(m6, test="LR")

dfwork$newparty <- dfwork$party 
levels(dfwork$party)
levels(dfwork$newparty) <- c("weak", "ind", "weak", "strong", "strong", "weak", "weak")

m8 <- glm(pres ~ poly(age,2)*c.edu + inter + close + sat + newparty, family=binomial, data=dfwork)
summary(m8)

#The models are nested therefore im allowed to use a deviance test anova
anova(m8, m7, test="Chisq")
waldtest(m8, m7, test="Chisq") #This is not going to work given the definition of my criteria (Wald test does not know that this models are in fact nested)
```

## Interactions

```{r}
#DO NOT DO THIS IN ASSIGNMENT 2!
m9 <- glm(pres ~ poly(age,2)*c.edu + poly(age,2) * (inter+close+sat+newparty)^2, family=binomial, data=dfwork)
m10 <- step(m9)
m11 <- step(m9, k=log(nrow(dfwork)))

summary(m10)
summary(m11)
anova(m10, m11, test="Chisq") # Both models are borderline equivalent...
```

# Model Validation and Diagnostics

```{r}
#Change for m11
#Ðo not use for GLMs 
plot(m7)

# Use for glms 
residualPlots(m7)
influencePlot(m7, id.n=5)

# 913, 951 can be considered influent data (However, this doesn't equal to being outlier or noise... no need to remove values with large hat values)
Boxplot(cooks.distance(m7), id=list(labels=rownames(dfwork)))

ids_rem <- which(rownames(dfwork) %in% c("913", "951")); ids_rem

#If we recalculate the ids_rem we get 2127
dfwork <- dfwork[-ids_rem, ]

```
