---
title: "Assingment1"
author: "Walter and Joan and Anna"
date: "`r Sys.Date()`"
output: html_document
---

# Cancer Mortality Analysis

## Description 

This project aims to analyse data from the famous "Cancer Mortality" open data set which contains over 3000 observations and 33 variables, included the target variable **target deathrate**.

We will follow a common data analysis pipeline:

```
Ingestion -> Cleaning -> Transformations -> EDA -> Feature Engineering -> Model selection -> evaluation
```

Credits:
- Anna Monso
- Joan Acero
- Walter Troiani

## 1. Data Ingestion 

```{r}
library(mice)
library(missMDA)
library(car)
setwd("~/Escritorio/College/Q9/SIM/ASSIGNMENT1")
test_df <- read.csv("test.csv",)
train_df <- read.csv("train.csv")

names(train_df)
```

## 2. Variable analysis 


```{r}
head(train_df)

```

*Observations* refer to huge populations of people (100K) that reside in same U.S county. 


The main *variables* we can see can be grouped in diferent parts. 

- **Demographical Data**:
  - *Ages*: 
    - `medianage` (Median age of the population)
    - `medianagemale` (Median age of males)
    - `medianagefemale` (Median age of females)
  - *Population*: 
    - `popest2015` (Population estimate for 2015)
    - `birthrate` (Birth rate)
  - *Marital Status*: 
    - `percentmarried` (Percentage of individuals married)
    - `pctmarriedhouseholds` (Percentage of married households)
  - *Racial Demographics*: 
    - `pctwhite` (Percentage of White population)
    - `pctblack` (Percentage of Black population)
    - `pctasian` (Percentage of Asian population)
    - `pctotherrace` (Percentage of Other Races)

- **Economic Data**:
  - *Income*: 
    - `medincome` (Median household income)
  - *Poverty*: 
    - `povertypercent` (Percentage of population below the poverty line)
  - *Study Spending*: 
    - `studypercap` (Per capita study spending)
  - *Binned Income*: 
    - `binnedinc` (Income categorized into bins)

- **Geographical Data**:
  - `geography` (Represents U.S. counties)

- **Cancer Statistics**:
  - `avganncount` (Average annual cancer cases)
  - `avgdeathsperyear` (Average cancer-related deaths per year)
  - `target_deathrate` (Targeted death rate per 100,000 people)
  - `incidencerate` (Cancer incidence rate per 100,000 people)

- **Educational Data**:
  - *Education for Age 18-24*:
    - `pctnohs18_24` (Percentage with no high school education)
    - `pcths18_24` (Percentage with a high school diploma)
    - `pctsomecol18_24` (Percentage with some college education)
    - `pctbachdeg18_24` (Percentage with a bachelor's degree)
  - *Education for Age 25 and Over*:
    - `pcths25_over` (Percentage with a high school diploma)
    - `pctbachdeg25_over` (Percentage with a bachelor's degree)

- **Employment Data**:
  - `pctemployed16_over` (Percentage of population aged 16+ employed)
  - `pctunemployed16_over` (Percentage of population aged 16+ unemployed)

- **Healthcare Coverage**:
  - *Private Insurance*:
    - `pctprivatecoverage` (Percentage with any private coverage)
    - `pctprivatecoveragealone` (Percentage with private coverage only)
    - `pctempprivcoverage` (Percentage of employed population with private coverage)
  - *Public Insurance*:
    - `pctpubliccoverage` (Percentage with any public coverage)
    - `pctpubliccoveragealone` (Percentage relying solely on public coverage)
    
    
```{r}

#train_df[1, c("pctwhite", "pctblack", "pctemployed16_over", "geography")]
hist(train_df$pctwhite)

```

## 3. Data Cleaning

```{r}
summary(train_df)
```

### 3.1 Missing values & Imputation
```{r}
md.pattern(x=train_df, rotate.names=TRUE)
```

### 3.2 Outlier detection & handling


```{r}

```

## 4. Data Transformations 

```{r}

```

## 5. Exploratory Data Analysis

```{r}

```

## 6. Modeling

We are going to implement various linear models to predict the response variable: **target_deathrate**. 

k-fold cross validation where k = 10 
:)

```{r}

```


## 7. Evaluation

```{r}

```

## 8. Assesing final questions 

Finally we need to asses some relevant statistical questions:

- Multicollinearity 
- Normality of residual distribution
- Linearity of the model (Parameters)
- Serial independence of errors
- Heteroskedasticity