---
title: 'Assignment1: CancerMortality'
author: "Joan Acero, Anna Monsó, Walter Troiani"
date: "2024-10-21"
output:
 pdf_document:
   toc: true
editor_options: 
  chunk_output_type: console
---

**HINT: Retain all available numeric variables except those that are not useful in explaining the target.**
# Questions: 

- Should we remove the outliers, apply transformations, etc.in the first analysis of the variables (Variable Analysis) ?


\newpage 
 

# Load the dataset and required libraries
```{r}

rm(list=ls())
knitr::opts_chunk$set(fig.align='center', out.width='50%') # Manage output plots width and align

train <- read.csv("train.csv")
test <- read.csv("test.csv")

library(moments)
library(car)

```

# Define some functions to avoid repetition of same processes
```{r}

calculate_outliers <- function(data) {
  # Check for regular outliers
  outliers <- boxplot.stats(data)$out
  num_outliers <- length(outliers)

  # Calculate the lower and upper bounds for severe outliers
  Q1 <- quantile(data, 0.25)
  Q3 <- quantile(data, 0.75)
  IQR <- Q3 - Q1

  lower_severe_bound <- Q1 - 3 * IQR
  upper_severe_bound <- Q3 + 3 * IQR

  # Identify severe outliers
  severe_outliers <- data[data < lower_severe_bound | data > upper_severe_bound]
  num_severe_outliers <- length(severe_outliers)

  # Return the results as a list
  return(list(
    num_outliers = num_outliers,
    num_severe_outliers = num_severe_outliers
  ))
}

```

\newpage

# Data Preparation

In this section, a descriptive analysis is conducted on each variable, a data quality report is generated, and imputation and profiling are addressed.


## Univariate Descriptive Analysis (Variable Analysis)

- Original numeric variables corresponding to qualitative concepts have to be converted to factors:
- Original numeric variables corresponding to real quantitative concepts are kept as numeric but additional factors should also be created
- Exploratory Data Analysis for each variable (numeric summary and graphic support) of the train sample.

### Variable *avganncount*

*avganncount* is a numerical variable representing the "Mean number of reported cancer cases diagnosed annually (2010-2015)." The difference between the mean (623.2) and the median (175) indicates that the distribution is not symmetric and deviates from normality. Additionally, the first and third quartiles (80 and 509, respectively) are much lower than the mean of 623.2 and the maximum value of 38150.0, further suggesting skewness. The boxplot confirms this skewness, with an extended upper whisker and numerous outliers, indicating a heavy right tail. It has 273 outliers, and 224 of these are severe outliers.

The histogram overlaid with a normal curve highlights the deviation from a normal distribution with the same mean and standard deviation. This variable has no missing values. Lastly, the skeweness indicator and Shapiro test rejection of the null hypothesis also confirm the non-normal distribution of the variable. 

```{r}

summary <- summary(train$avganncount); summary
mm <- mean(train$avganncount)
ss <- sd(train$avganncount)
range <- (summary[[6]] - summary[[1]]); range
boxplot(train$avganncount, main= "Boxplot")
hist(train$avganncount,freq = F,30, main = "Histogram")
curve(dnorm(x, mean = mm, sd = ss),col="red", lwd = 2, add=T)
skewness(train$avganncount)
shapiro.test(train$avganncount)

calculate_outliers(train$avganncount)


```

### Variable *avgdeathsperyear*

In a similar way, *avgdeathsperyear* is a numerical variable expressing the "Mean number of reported mortalities due to cancer". It has no missing values. The difference between the mean (191.6) and median (62) indicates that the distribution is not symmetric. Moreover, the first and third quartiles (29 and 140.5, respectivelly) are lower than the mean and the maximum value of 14010, further suggesting a skewed ditribution. With the help of graphical tools like the boxplot we can confirm the skeweness, obtaining 225 outliers (138 of which are severe), indicating a large right tail. The histogram overlaid with a normal curve highlights the deviation from a normal distribution with the same mean and standard deviation, and the Shapiro test also confirms the non-normality of the variable. 

```{r}

variable <- train$avgdeathsperyear;
summary <- summary(variable);summary
mm <- mean(variable)
ss <- sd(variable)
range <- (summary[[6]] - summary[[1]]); range
boxplot(variable, main = "Boxplot")
hist(variable,freq = F,100, main = "Histogram")
curve(dnorm(x, mean = mm, sd = ss),col="red", lwd = 2, add=T)
skewness(variable)
shapiro.test(variable)

calculate_outliers(variable)

```

### Variable *target_deathrate* :

Variable *target_deathrate* is a numerical variable that is the target of the model. It contains no missing values in the train sample. The median (178.3) and mean (178.8) values are really close, expressing the symmetry of the distribution of the variable. Thanks to the boxplot and the histogram we can graphically confirm the symmetry of the distribution. 

We performed the Shapiro test to check for the normality of the data, and obtained a small p-value that rejects the null hypothesis, indicating that the variable does not follow a normal distribution. Therefore, the target variable is not normally distributed. 

Moreover, it contains 35 outliers, 1 of which is a severe.  

```{r}

variable <- train$target_deathrate;
summary <- summary(variable); summary
mm <- mean(variable)
ss <- sd(variable)
range <- (summary[[6]] - summary[[1]]); range
boxplot(variable, main = "Boxplot")
hist(variable,freq = F,100, main = "Histogram")
curve(dnorm(x, mean = mm, sd = ss),col="red", lwd = 2, add=T)


skewness(variable)
shapiro.test(variable)

calculate_outliers(variable)

```

### Variable *incidencerate* : 

Variable *incidencerate* is a numerical variable that expresses "Mean per capita (100,000) cancer diagnoses". The variable has no missing values. It has similar mean (449) and median (453.5) values given its range and suggesting a simmetry distribution. It contains 60 outliers, 7 of which are severe outliers. In the boxplot we can see how the most severe outliers are high and not low values, corresponding to Osceola County, Florida and Williamsburg city, Virginia. The histogram with the overlapping normal distribution shows a quite symmetrical distribution, but the Shapiro test and skeweness value discard the normality of the variable's distribution.

```{r}

variable <- train$incidencerate;
summary <- summary(variable); summary
mm <- mean(variable)
ss <- sd(variable)
range <- (summary[[6]] - summary[[1]]); range
Boxplot(variable, main = "Boxplot")
train[889,"geography"]
train[168, "geography"]
hist(variable,freq = F,100, main = "Histogram")
curve(dnorm(x, mean = mm, sd = ss),col="red", lwd = 2, add=T)


skewness(variable)
shapiro.test(variable)

calculate_outliers(variable)

```

### Variable *medincome* :

Variable *medincome* is a numerical variable that expresses the "Median income per county". Thanks to the histogram, Shapiro test, and skeweness value, we see how the distribution is neither Normal nor symmetric, but it is skewed with a long right tail. It has 69 outliers, 15 of them severe. The highest value is for Loudoun County, Virginia. There are no missing values in this variable. 

```{r}

variable <- train$medincome;
summary <- summary(variable); summary
mm <- mean(variable)
ss <- sd(variable)
range <- (summary[[6]] - summary[[1]]); range
Boxplot(variable, main = "Boxplot")
train$geography[128] # Virginia
hist(variable,freq = F,100, main = "Histogram")
curve(dnorm(x, mean = mm, sd = ss),col="red", lwd = 2, add=T)


skewness(variable)
shapiro.test(variable)
calculate_outliers(variable)

```

### Variable *popest2015* :

Variable *popest2015* is a numerical variable that expresses the population of the county. It does not count with missing values. It has a long range and is highly skewed, with a long right tail. In fact, it contains 252 outliers, 162 of which are severe, and all of them are above the upper whisker in the boxplot. The highest value is for Los Angeles County, California. 
The large difference between the mean (106.841) and median (27.158), along with the spread between the 1st (12.191) and 3rd quartiles (66.880), indicates a positively skewed distribution, where extreme high values are pulling the mean upward.

```{r}

variable <- train$popest2015;
summary <- summary(variable); summary
mm <- mean(variable)
ss <- sd(variable)
range <- (summary[[6]] - summary[[1]]); range
Boxplot(variable, main = "Boxplot")
train$geography[600]
hist(variable,freq = F,100, main = "Histogram")
curve(dnorm(x, mean = mm, sd = ss),col="red", lwd = 2, add=T)

skewness(variable)
shapiro.test(variable)
calculate_outliers(variable)

```



### Variable *povertypercent* : 


Variable *povertypercent* is a numerical variable that expresses the percent of population in poverty and has no missing values. Similarly to other variables, it has a skewed distribution with a larger right tail. It has 42 outliers, but none of them is severe. The p-value obtained in the shapiro_test confirms the rejection of the null hypothesis, concluding that the variable does not have a normal distribution. 

```{r}

variable <- train$povertypercent;
summary <- summary(variable); summary
mm <- mean(variable)
ss <- sd(variable)
range <- (summary[[6]] - summary[[1]]); range
Boxplot(variable, main = "Boxplot")
hist(variable,freq = F,100, main = "Histogram")
curve(dnorm(x, mean = mm, sd = ss),col="red", lwd = 2, add=T)

skewness(variable)
shapiro.test(variable)
calculate_outliers(variable)

```


### Variable *studypercap* : I CONSIDER IT IS WORTH TO REMOVE


Variable *studypercap* is a numerical variable that expresses "Per capita number of cancer-related clinical trials per county". It contains no missing values, and has 307 outliers (222 severe). Its distribution is non-symmetric. Note that 1162 out of 1831 observations contain value 0 in this variable, which indicates that it may only provide noise to the model. **Thus, we consider it is worth to remove the variable. **


```{r}

variable <- train$studypercap;
summary <- summary(variable); summary
length(variable[variable==0])
mm <- mean(variable)
ss <- sd(variable)
range <- (summary[[6]] - summary[[1]]); range
Boxplot(variable, main = "Boxplot")
hist(variable,freq = F,100, main = "Histogram")
curve(dnorm(x, mean = mm, sd = ss),col="red", lwd = 2, add=T)

calculate_outliers(variable)

```


### Variable *binnedi* : I CONSIDER IT IS WORTH TO REMOVE


Variable *binnedi* is a categorical variable (or factor) that expresses the median income per capita binned by decile. The variable 

-- anna -- I'll remove the variable if we require precision or need the full detail of income differences but
if we want to compare groups maybe is useful to keep it. 


```{r}

variable <- train$binnedinc;
summary <- summary(variable); summary
table(variable)
prop.table(table(variable))
barplot(table(variable), main="Bar Plot of Variable", xlab="Categories", ylab="Frequency")

```


### Variable *medianage* :
Medianage is a numerical variable that expresses the median age of county residents. The variable has no missing values. Observing its mean(45.25), the median (40.90) and the 3 Qu. (44.90) being lower than the mean suggest skewness. Has 65 outliers, 18 of which are severe outliers. The highest value is in "Iosco County, Michigan". After anylizing the boxplot, and the histogram along the Shapiro test and skeweness value we can discard the normality of the variable's distribution.


```{r}
variable <- train$medianage
summary <- summary(variable) ; summary
sum(is.nan(variable))
mm <- mean(variable)
ss <- sd(variable)
range <- (summary[[6]] - summary[[1]]); range
Boxplot(variable, main = "Boxplot")
train$geography[193]
hist(variable,freq = F,100, main = "Histogram")
curve(dnorm(x, mean = mm, sd = ss),col="red", lwd = 2, add=T)

calculate_outliers(variable)
shapiro.test(variable)

```


### Variable *medianagemale* :
Numeric variable, represents the mean age of the male population of each county. There are no missing values and it has 46 outliers, non of them several outliers. The mean is 39.59 yo and the median 39.50yo what suggest a simmetry distribution. After looking at the results of the shapiro-test it doesn't have a normal distribution 

```{r}
variable <- train$medianagemale
summary <- summary(variable) ; summary
sum(is.nan(variable))
mm <- mean(variable)
ss <- sd(variable)
range <- (summary[[6]] - summary[[1]]); range
Boxplot(variable, main = "Boxplot")

hist(variable,freq = F,100, main = "Histogram")
curve(dnorm(x, mean = mm, sd = ss),col="red", lwd = 2, add=T)

calculate_outliers(variable)
shapiro.test(variable)

```

### Variable *medianagefemale* : I AM NOT SURE IF IT'S WORTH IT TO SEPARATE THE MEDIANAGE WITH SEX

The medianagefemale is a numerical variable that analyze the median age of the females in each county. It has a simmetric distribution (median=42.40, mean=42,17) but after analyzing the results of the Shapiro-test we observe that does not have a normal distribution. 

-- comment --
!! I think that something it's not right because the mean of males is 39.59 yo, the mean of females is 42,17 yo, therefore the mean of the medianage should be around 40,88 yo, but it's 45.25yo

```{r}
variable <- train$medianagefemale
summary <- summary(variable) ; summary
sum(is.nan(variable))
mm <- mean(variable)
ss <- sd(variable)
range <- (summary[[6]] - summary[[1]]); range
Boxplot(variable, main = "Boxplot")

hist(variable,freq = F,100, main = "Histogram")
curve(dnorm(x, mean = mm, sd = ss),col="red", lwd = 2, add=T)

calculate_outliers(variable)
shapiro.test(variable)
```


### Variable *geography* :
Categorical variable that shows the name of each county. We can use this variable as a PK 

--comment -- 
Maybe we can separate the name of the state and create a new varible to be able to study the distribution around the United States 


```{r}
variable <- train$geography;
summary <- summary(variable); summary
table(variable)

```

### Variable *percentmarried* 
Numerical variable that represent the percent of county residents who are married. It has a non-symmetric distribution, it is skewed with a long left tail observed in the histogram. It doesn't have missing values, it has 34 outliers (non of them severe) and they are concentrated on the lower end of the distribution. With the shapiro-test we can confirm the rejection of the null hypothesis, therefore it doesn't have a normal distribution.   


```{r}
variable <- train$percentmarried
summary <- summary(variable) ; summary
sum(is.nan(variable))
mm <- mean(variable)
ss <- sd(variable)
range <- (summary[[6]] - summary[[1]]); range
Boxplot(variable, main = "Boxplot")

hist(variable,freq = F,100, main = "Histogram")
curve(dnorm(x, mean = mm, sd = ss),col="red", lwd = 2, add=T)

calculate_outliers(variable)
shapiro.test(variable)

```

### Variable *pctnohs18_24* 

Numerical variable that show the percentage of county residents aged 18-24 whose highest level of education is less than a high school diploma. The distribution of this variable is skewed with a long right tail, we can confirm that it doesn't have a normal distribution thanks to the histogram and the curve. Moreover the p-value obtained in the shapiro-test confirm the non normal distribution. It doesn't have misssing values, and it has 35 outliers, 5 of them severe. 

```{r}

variable <- train$pctnohs18_24
summary <- summary(variable); summary
mm <- mean(variable)
ss <- sd(variable)

Boxplot(variable, main="Boxplot")
hist(variable , freq=F, 100, main="Histogram")
curve(dnorm(x,mean=mm,sd=ss),col='red', lwd=2, add = T)

calculate_outliers(variable)
shapiro.test(variable)

```

### Variable *ppcths18_24* 
Numeric variable, it represent the percentage of county residents ages 18-24 highest education attained is a high school diploma. We can observe with the median(34,7%) and the mean(35%) that the variable has a symmetric distribuition, but after the shapiro_test we observe that it doesn't have a normal distribution. It doesn't have missing values and it has 33 outliers, non of them severe. 
```{r}
variable <- train$pcths18_24

summary <- summary(variable); summary
mm <- mean(variable)
ss <- sd(variable)

Boxplot(variable, main="Boxplot")
hist(variable , freq=F, 100, main="Histogram")
curve(dnorm(x,mean=mm,sd=ss),col='red', lwd=2, add = T)

calculate_outliers(variable)
shapiro.test(variable)

```

### Variable *pctsomecol18_24* NOT WORTH TO USE IT (75% MISSING VALUES)

Numerical variable. Percent of county residents ages 18-24 highest education attained some college. It has 1375 missing values. (75%) therefore we would remove this variable. 


```{r}
variable <- train$pctsomecol18_24
summary <- summary(variable); summary

```

### Variable pcths25_over*
Numerical variable that express the percentage of county residents ages 25 and over highest education attained a high school diploma. It doesn't have missing values but it have 18 outliers, non of the severe. It doesn't have a normal distribution.
```{r}
variable <- train$pcths25_over
summary <- summary(variable); summary
mm <- mean(variable)
ss <- sd(variable)

Boxplot(variable, main="Boxplot")
hist(variable , freq=F, 100, main="Histogram")
curve(dnorm(x,mean=mm,sd=ss),col='red', lwd=2, add = T)

calculate_outliers(variable)
shapiro.test(variable)
```














## Data Quality Report 

Per variable, count:

- Number of missing values.
- Number of errors (including inconsistencies).
- Number of outlines.
-Rank variables according the sum of missing values (and errors).


```{r}

# Number of missing values.
# Number of errors (including inconsistencies).
# Number of outlines.
# Rank variables according the sum of missing values (and errors).

```


Per individuals, count:

- Number of missing values
- Number of errors
- Number of outliers
- Identify individuals considered as multivariant outliers.


```{r}

#Number of missing values
#Number of errors
#Number of outliers
#Identify individuals considered as multivariant outliers.

```

Create a variable adding the total number of missing values, outliers and errors. Describe these variables, to which other variables exist higher associations.

- Compute the correlation with all other variables. Rank these variables according the correlation
- Compute for every group of individuals the mean of missing/outliers/error values. Rank those groups according the computed mean. 

```{r}

#Include Code

```


## Imputation

- Numeric Variables
- Factors

```{r}

#Include Code

```


## Profiling

- Target : Death rate

*Model fitting to train sample only!!*


```{r}

#Include Code

```







