---
title: 'Assignment1: CancerMortality'
author: "Joan Acero, Anna Monsó, Walter Troiani"
date: "2024-10-21"
output:
 pdf_document:
   toc: true
---


\newpage

HINT: Retain all available numeric variables except those that are not useful in explaining the target.


# Load the dataset
```{r}

rm(list=ls())
train <- read.csv("train.csv")
test <- read.csv("test.csv")


```

# Data Preparation

## Univariate Descriptive Analysis (Variable Analysis)

For each variable: 

- Original numeric variables corresponding to qualitative concepts have to be converted to factors

```{r}

# Include Code


```

- Original numeric variables corresponding to real quantitative concepts are kept as numeric but additional factors should also be created

```{r}

#Include Code

```

- Exploratory Data Analysis for each variable (numeric summary and graphic support).

```{r}

#Include Code

```


## Data Quality Report 

Per variable, count:

- Number of missing values
- Number of errors (including inconsistencies)
- Number of outliers
-Rank variables according the sum of missing values (and errors)

```{r}

#Include Code

```


Per individuals, count:

- Number of missing values
- Number of errors
- Number of outliers
- Identify individuals considered as multivariant outliers.

```{r}

#Include Code

```

Create a variable adding the total number of missing values, outliers and errors. Describe these variables, to which other variables exist higher associations.

- Compute the correlation with all other variables. Rank these variables according the correlation
- Compute for every group of individuals the mean of missing/outliers/error values. Rank those groups according the computed mean. 

```{r}

#Include Code

```


# Imputation

- Numeric Variables
- Factors

```{r}

#Include Code

```


# Profiling

- Target : Death rate


```{r}

#Include Code

```







