---
title: "Lab1-Session2"
author: "Walter J.T.V"
date: "`r Sys.Date()`"
output: html_document
---
# Lab 1 - Session2 - Prestige Imputation
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
```{r}

library(car)
library(missMDA) #FactoMineR
library(mice)

Prestige <- Prestige
df <- Prestige
df$census <- NULL

#Define a binary factor variable "f.prof":  (Yes, Prof), (No, Rest) if an observation is a professor
df$f.prof <- 0
df$f.prof[df$type =="prof"] <- 1
df$f.prof <- factor(df$f.prof, labels=c("No", "Yes"))
summary(df)
```
## Imputation
### Numeric variables
Artificially inducing missing data just for demonstration purposes. 
The imputation is never done by the mean, therefore we are going to do imputation using PCA numeric variables. That will make the imputed value multidimensionally coherent with all the other variables (Multivariate properties). 
```{r}
#Assume women==0 is missing data
ll <- which(df$women == 0)
df$women[ll] <- NA
summary(df)

# imputePCA() --> numeric vars
# imputeMCA() --> factor vars

res.impPCA <- imputePCA(df[, 1:4])
summary(res.impPCA$completeObs)
```
This PCA could land extreme values, negative ones or others that might not make sense at all, so we are going to check the difference in distributions from the original data vs the one that has recieved imputation. Taking a look into their quantiles or histogram would be enough

```{r}

q1 <- quantile(df$women, prob=seq(0, 1, 0.1), na.rm=T)
q2 <- quantile(res.impPCA$completeObs[,3], prob=seq(0,1,0.1))
```
Note: completeObs is a data-frame containing the result
Alternatively, we could perform the imputation using the MICE library 

```{r}
?mice
res.mice <- mice(df)
summary(complete(res.mice))
```
### Categorical variables

```{r}

df$type <- Prestige$type 
res.impMCA <- imputeMCA(df[c("type", "f.prof")])

#Before and after
summary(df[c("type", "f.prof")])
summary(res.impMCA$completeObs)
```