---
title: "ordinal_polytomous_models"
author: "Walter J.T.V"
date: "2024-12-16"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Ordinal modelling

In this lab 12 we are going to use polytomous ordinal models to predict

```{r}
library(MASS)
library(car)
library(effects)
rm(list=ls())
load("Fesal1985.RData")
summary(fesal85)
df <- fesal85
```


## Nominal Model

This model has 3 groups, therefore 2 equations:

$$ log\frac{\pi_{other}}{\pi_{none}} = \eta_1$$

$$ log\frac{\pi_{ster}}{\pi_{none}} = \eta_1$$


```{r}
#Final model of past lab 11
mm4a <- multinom(c.use ~ I(n.age - 22) + I((n.age-22)^2) + I((n.age-22)^3), weights=y, data=fesal85)
summary(mm4a)
stepAIC(mm4a)
```

## Hierarchical Logit Model

This model with 3 values has 2 hierarchical levels, therefore 2 equations:
      
          H1         H2
root ----> None 
     |---> Some ----> Other 
                |---> Ster

```{r}
fesal85$bc.use <- factor(ifelse(fesal85$c.use=="None",0,1), labels=c("None", "Some"))
table(fesal85$bc.use)
h1 <- glm(bc.use ~ I(n.age - 22) + I((n.age-22)^2), family=binomial, data=fesal85, weight=y)
summary(h1)
h2 <- glm(c.use ~ I(n.age-22) + I((n.age - 22)^2) + I((n.age - 22)^3), family=binomial, data=fesal85[fesal85$bc.use == "Some", ], weight=y)
summary(h2)
AIC(h1) + AIC(h2) > AIC(mm4a)
```

## Ordinal Model (Today is the day!)

### Simple model logit
This model will produce 2 equations also with a latent scale. Ordinal models have cumulative probabilities, so summing up all probablities must sum up to 1.

$$ \gamma_{none} = \pi_{none} |||  \gamma_{other} = \pi_{none} + \pi_{other} ||| \gamma_{ster}  = \pi_{none} + \pi_{other} + \pi_{ster} = 1$$
$$ logit(\gamma_{j}) = \{\eta_{none}, \eta_{other}\}$$
```{r}

table(fesal85$c.use, useNA="always")
ord0 <- polr(c.use ~ 1, weights=y, data=fesal85)
summary(ord0)

ord0$zeta
```

In this case, the logit values of the intercepts are (This is a simple model with just intercepts):

$$ logit(\gamma_{none}) = 0.112 ||| logit(\gamma_{other}) = 0.7651$$
Therefore the raw probablities is just the sigmoid function of the logit(gamma):

$$\gamma_{none} = \frac{e^{0.112}}{1 + e^{0.112}} = 0.1119... ||| \gamma_{other} = \frac{e^{0.7651}}{1 + e^{0.7561}} = 0.76512...$$

### Simple model probit

Now lets play with probit link function, that uses the normal distribution:

$$ F^{-1}_z() = qnorm \implies F_z = pnorm$$

$$ probit(\gamma_j) = \eta_j \implies \gamma_j = F_z(\eta_j)$$
```{r}

ord0p <- polr(c.use ~ 1, weights=y, data=fesal85, method="probit")
summary(ord0p)
probit_vals <- ord0p$zeta

probability_distrib <- pnorm(probit_vals)

diff(c(0, pnorm(ord0p$zeta), 1))
```

### Ordinal model using explanatory variables

```{r}
#logit to the cumulative probability 
ord1f <- polr(c.use ~ g.age, weights=y, data=fesal85)
summary(ord1f)

ord1 <- polr(c.use ~ n.age, weights=y, data=fesal85)
summary(ord1)

#We cannot use deviance inferential tests given that the 2 models aren't nested 
#However this motherfucker will tank the shit out of the input and won't complain so be careful lil bro
anova(ord1f, ord1, test="Chisq") #NOT CORRECT!!!!
#However we can use Akaike or Bayesian information criteria (AIC/BIC) to decide which of the 2 models is likely to be better
AIC(ord1, ord1f)


#We can take a look at the t-values given that the polynomials are orthogonal
ord3 <- polr(c.use ~ poly(n.age, 3), weights=y, data=fesal85)
summary(ord3)

ord2 <- polr(c.use ~ poly(n.age, 2), weights=y, data=fesal85)
summary(ord2)

#Big p-value, we cannot reject that the models are equivalent, therefore third order isn't needed
anova(ord2, ord3, test="Chisq")

#For better interpretability/explainability use I instead of Poly
ord2 <- polr(c.use ~ I(n.age) + I((n.age)^2), weights=y, data=fesal85)
summary(ord2)

#This are the predicted values, but in logit scale
etaNone <- 7.567 - 0.451065 * 22 + 0.006565 * (22^2)
etaOther <- 8.27 - 0.451065 * 22 + 0.006565 * (22^2)

#Now lets get the gamma probabilities (cumulative!!!) using the sigmoid function (0,1)
gamNone <- exp(etaNone) / (1 + exp(etaNone)); gamNone
gamOther <- exp(etaOther) / (1 + exp(etaOther)); gamOther

#Now lets get the raw probabilities (no cumulative)
piNone <- gamNone; piNone
piOther <- gamOther - piNone; piOther


#Verify that predictions match our hand-made calculations with R
predict(ord2, newdata=data.frame(n.age=22), type="probs")
```
Signs must be reversed (Latent values are the inverse sign of the coefficient)

The 2 model equations are the following: $$logit(\gamma_j) = \{7.57, 8.27\} - 0.461\cdot n.age + 0.0065\cdot n.age^2$$

If we want to predict the values when n.age = 22 we can either use **el truquito** or just replace the values in the predictor variables (X):  (LOGIT SCALE!)

$$logit(\gamma_{none}) = 7.57 - 0.461\cdot 22 + 0.0065\cdot 22^2 = \eta_{none} = $$
$$logit(\gamma_{otther}) = 8.27 - 0.461\cdot 22 + 0.0065\cdot 22^2 = \eta_{other} = $$

Then its clear that what changes between these models is intercept, because in this case we share coefficients.

### Model comparison (Ordinal vs Nominal vs Hierarchical)

```{r}
#Hierachical AIC
AIC(h1) + AIC(h2)

#Nominal best model AIC <--- This is the best based on Akaike information Criteria
AIC(mm4a)

#Ordinal best model AIC
AIC(ord1f)

#This is what the model does, but not what the data does
plot(Effect(focal.predictors = c("n.age"), mm4a), rug=F, layout=c(3,1))
#This cannot be done for hierarchical models...
# ordinal model
plot(Effect(focal.predictors = c("g.age"), ord1f), rug=F, layout=c(3,1))

```