---
title: "polytomous_modeling"
author: "Walter J.T.V"
date: "2024-12-09"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Polytomous modeling (Softmax / Multinomial)

There are 3 kinds of polytomous models: 

1. Nominal:
2. Ordinal:
3. Hierarchical: 

## Data ingestion 

```{r}
library(car)
library(effects)
library(scales)
library(MNLpred)
library(MASS) # Ordinal models
library(nnet) # Nominal models
library(ggplot2)

load("Fesal1985.RData")
summary(fesal85)
df <- fesal85
```

Nominal: 3 levels (1 -> None, 2 -> Other, 3 -> Ster)

The weights are = positive outcome

j = 1/7 explain variable
(1+6) variables (7)
$$ log(\frac{\pi_{other}}{\pi_{none}}) = \eta_{other} + \alpha^{other}_j$$
(1+6) variables (7)
$$ log(\frac{\pi_{ster}}{\pi_{none}}) = \eta_{ster} + \alpha^{ster}_j$$
The model has 14 variables in total
The deviance of this model will be: 

$$ D(y, \hat{\mu}) = 2* \Sigma \Sigma (   o_{ij} * log(\frac{o_{ij}}{lol?})) = 6267$$

## Model exploration

### Null model

```{r}

nm0 <- multinom(c.use ~ 1, weights=y, data=df)
summary(nm0) # 6270 deviance, computed for all 3 categories (Formula in the slides)
```

### Model 1 (Categorical grouped age)
```{r}

?multinom
mm1 <- multinom(c.use ~ g.age, data=df, weights = y)
summary(mm1)

anova(nm0, mm1, test="Chisq")
```

This model has smaller deviance, in fact could be compared by means of inferential testing (Chi-squared distributed test anova test)

$$ \Delta D = 6271 - 5746 = 521...$$
$$ pvalue = P(\chi^2_{12} > 525) << 0.05 \implies Reject$$

### Model 2 (Numeric variable age)
```{r}

#We don't get inferential tests of \beta = 0, however we can anova test the shit out of them
mm2 <- multinom(c.use ~ n.age, weights=y, data=df)
mm3 <- multinom(c.use ~ poly(n.age,2), weights=y, data=df)
#These weights are in orthogonal weird ass scale, it is not log odds
mm4 <- multinom(c.use ~ poly(n.age,3) , weights=y, data=df)
summary(mm4)

anova(mm3, mm4, test="Chisq") # We reject that both models are equivalent

#This way we transform it to log odds scale
mm4m <-  multinom(c.use ~ I(n.age^2) + I(n.age^3) , weights=y, data=df)
```

$$ log\frac{\pi_{other}}{\pi_{none}} = -3.25 + 0.125n.age + 0.00012n.age^2 - 0.000056n.age^3 $$

### Predictions

We want to forecast at age 22 the probability distribution in the 3 plausible classes: We get a 64% None, 23% Other, 12% Ster.
```{r}

predict(mm4, newdata=data.frame(n.age=22), type="probs")

mm4_aux <- multinom(c.use ~ I(n.age-22) + I((n.age-22)^2) + I((n.age-22)^3), weights=y, data=df)
#Center the values at 22 ;) (It's our target prediction value, its just a trick to avoid computation with a fucking calculator)
summary(mm4_aux)
```


The trick :
$$ log\frac{\pi_{other}}{\pi_{none}} = -1.03 + 0.049 (22-22) - 0.0035(22-22)^2 -0.00005 (22-22)^3 $$

Now we can get the normal odds applying a simple transformation

$$ log\frac{\pi_{other}}{\pi_{none}} = -1.03... \implies \frac{\pi_{other}}{\pi_{none}} = e^{-1.03...} = 0.355$$
$$ log\frac{\pi_{ster}}{\pi_{none}} = -1.65... \implies \frac{\pi_{ster}}{\pi_{none}} = e^{-1.65...} = 0.192$$
$$\pi_{none} + \pi_{other} + \pi_{ster} = 1  \implies \pi_{none} = \frac{1}{1 + \frac{\pi_{other}}{\pi_{none}} + \frac{\pi_{ster}}{\pi_{none}} } = \frac{1}{1 + 0.355 + 0.192} $$

```{r}

first_odds <- exp(coef(mm4_aux)[1,1]) # First example look above dumbass
sec_odds <- exp(coef(mm4_aux)[2,1]) #Second example

prob_none <- 1/(1+first_odds + sec_odds); prob_none
prob_other <- prob_none * first_odds; prob_other
prob_ster <- prob_none * sec_odds; prob_ster
```

GLM Polytomous is kind of smelly due to the lack of residual analysis integrated in R, therefore we have to compute things manually

## Model comparison

Which model is better of the 2, that with numerical age or categorical grouped age:

```{r}
AIC(nm0, mm1, mm2, mm3, mm4)

#this is what the model effects are, not what the data does... Data distribution of each should be checked
plot(Effect(focal.predictors=c("n.age"), mm4 ), rug=FALSE, layout=c(3,1))
```


## Ordinal modeling

None < Other < Ster 

$$ \gamma_{none} = \pi_{none}; \gamma_{other} = \pi_{none} + \pi_{other}; \gamma_{star} = 1$$

## Hierarchical modeling 

This must be done by hierarchical levels (1,2,3...). Following up the hierarchical logit models:

Given that this is a hierarchy the probabilities are conditioned by the upper hierarchical levels, just like in probability trees in basic probability.

```{r}

#LEVEL 1
df$bc.use <- factor(ifelse(df$c.use=="None",0,1), labels=c("None", "Some"))
summary(df)
h1b3 <- glm(bc.use ~ poly(n.age, 3), family=binomial(), data=df) #Third order polynomial is not significant
summary(h1b3)

#Center the model to n.age=22
h1b2 <- glm(bc.use ~ I(n.age-22) + I((n.age-22)^2), family=binomial(), weights=y, data=df) #Third order polynomial is not significant
summary(h1b2)

#LEVEL 2
h2b3 <- glm(c.use ~  I(n.age-22) + I((n.age-22)^2) + I((n.age-22)^3),family=binomial(), weights=y, data=df[df$bc.use=="Some", ])
plot(allEffects(h2b3))
summary(h2b3)

prob_sterCONDITIONALTOsome <- predict(h1b3, newdata=data.frame(n.age=22), type="response") #Probability "Some" at n.age 22
prob_otherCONDITIONALTOsome <- 1 - predict(h2b3, newdata=data.frame(n.age=22), type="response") #Probabilty of "Other" at n.age 22
```

$$ \pi_{ster | some} = 0.35$$
$$ \pi_{other | some} = 0.65$$

$$ \pi_{other} = \pi_{some} * \pi_{other | some} $$

